---
title: "State-space models"
author: "Robin Aldridge-Sutton"
date: "04/04/2021"
output:
  github_document:
    toc: True
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Definition

A state-space model can be made up of:

- A state variable $x_t$, taking values in some state-space
- An output variable $y_t$
- An observation equation $P(y_t | x_t)$
- A state transition equation $P(x_{t + 1} | x_t)$
- Where $x_t$ and $y_t$ satisfy the Markov equations: 
$$P(y_{t + 1}, x_{t + 1} | y_t, x_t, ..., y_1, x_1) = P(y_{t + 1}, x_{t + 1} | y_t, x_t)$$
$$P(y_{t + 1} | x_{t + 1}, y_t, x_t) = P(y_{t + 1} | x_{t + 1})$$
- Which implies that $P(y_{t + 1}, x_{t + 1} | y_t, x_t) = P(y_{t + 1} | x_{t + 1}) P(x_{t + 1} | x_t)$

Obtaining the conditional distribution $P(x_t | y_t, ..., y_1)$, from which $P(x_{t + 1})$ and $P(y_{t + 1})$ can be derived/sampled, is known as filtering.

### White noise

White noise is the core of a stochastic time series model.

$$e_t \sim N(0, \sigma^2)$$

```{r}
# Set simulation parameters
n_t <- 20
t <- 1:n_t
n_reals <- 5
n_samps <- n_t * n_reals
quants <- t(matrix(c(0, 1.96, -1.96), nrow = 3, ncol = n_t))
q_lty <- c(1, 2, 2)

# Simulate white noise
e <- matrix(rnorm(n_samps), nrow = n_t)

# Plot it
matplot(t, e, type = 'l', main = "White noise", col = 1, lty = 1)
matlines(t, quants, col = 'red', lty = q_lty)
```

### Random walk

The steps in a random walk are white noise.

$$a_t \sim N(0, \tau^2)$$
$$\mu_{t + 1} = \mu_t + a_t$$
$$\mu_{t + 1} | \mu_t \sim N(\mu_t, \tau^2)$$
$$u_0 = 0 \implies \mu_t \sim N(0, t \tau^2)$$

```{r}
# Simulate random walk 
a <- matrix(rnorm(n_samps), nrow = n_t)
mu <- apply(a, 2, cumsum)

# Plot it
matplot(t, mu, type = 'l', main = "Random walk", col = 1, lty = 1)
matlines(t, quants * sqrt(t), col = 'red', lty = q_lty)
```

### Quarterly random walk

$$\mu_{t + 1} = \mu_{t - 3} + a_t$$
$$\mu_{t + 1} | \mu_{t - 3} \sim N \left( \mu_{t - 3}, \tau^2 \right)$$
$$u_0 = 0 \implies \mu_t \sim N \left( 0, ceiling \left( \frac{t}{4} \right) \tau^2 \right)$$

```{r}
# Simulate quarterly random walk 
mu_q <- apply(a, 2, function(column) 
  as.vector(t(apply(matrix(column, ncol = 4, byrow = T), 2, cumsum))))

# Plot it
matplot(mu_q, type = 'l', xlab = "Time", ylab = "Value",
        main = "Quarterly random walk", col = 1, lty = 1)
grid(nx = n_t / 4, ny = NA)
matlines(t, quants * sqrt(ceiling(t / 4)), col = 'red', lty = q_lty)
```

### Evolving variance

$$h_{t + 1} = h_t + a_t$$
$$y_t = \exp \left( \frac{h_t}{2} \right) e_t$$
$$y_t \not\sim N$$
$$y_t | h_t \sim N(0, \exp(h_t))$$
$$h_0 = 0 \implies \log\{var(y_t)\} \sim N(0, t \tau^2)$$

```{r}
# Simulate evolvng variance
h <- mu
y_v <- exp(h / 2) * e

# Plot it
matplot(y_v, type = 'l', xlab = "Time", ylab = "Value",
        main = "Evolving variance", col = 1, lty = 1)
lines(rep(0, n_t), col = 'red')
```

### Filtering for a random walk plus white noise model

$$y_t = \mu_t + e_t$$
$$y_t | \mu_t \sim N(\mu_t, \sigma^2)$$
$$y_t \sim N(0, t \tau^2 + \sigma^2)$$

```{r}
# Simulate and plot random walk plus white noise
y <- mu + e
matplot(t, y, type = 'l', main = "Random walk plus white noise", col = 1, 
        lty = 1)
matlines(t, quants * sqrt(t + 1), col = 'red', lty = q_lty)
```
$$x_t := \mu_t$$
$$\tilde{y}_t \sim N(\tilde{x}_t, \sigma^2 I_t)$$
$$\tilde{x}_t \sim N(\tilde{x}_{t - 1}, \tau^2 I_t)$$
$$\implies (\tilde{x}_t, \tilde{y}_t) \sim N$$
$$E(\tilde{x}_t | \tilde{y}_t) = E(\tilde{x}_t) + Cov(\tilde{x}_t, \tilde{y}_t) Var(\tilde{y}_t)^{-1} \{ \tilde{y}_t - E(\tilde{y}_t) \}$$
$$Var(\tilde{x}_t | \tilde{y}_t) = Var(\tilde{x}_t) - Cov(\tilde{x}_t, \tilde{y}_t) Var(\tilde{y}_t)^{-1} Cov(\tilde{y}_t, \tilde{x}_t)$$
$$E(\tilde{x}_t) = E(\tilde{y}_t) = 0$$
$$Cov(\tilde{x}_t, \tilde{y}_t) = Cov(\tilde{y}_t, \tilde{x}_t)' = (Cov(\tilde{x}_t, \tilde{x}_t) + Cov(\tilde{e}_t, \tilde{x}_t))' = Var(\tilde{x}_t)$$
$$Var(\tilde{y}_t) = Var(\tilde{x}_t) + \sigma^2 I_t$$
$$
\begin{bmatrix} 
1  & &  &\\ 
-1 & 1 & & \\ 
& ... & \\ 
 & & -1 & 1 \\ 
\end{bmatrix}
\begin{bmatrix} 
x_1 \\ 
... \\ 
x_t \\ 
\end{bmatrix} = 
\begin{bmatrix} 
1 \\ 
 \\ 
 \\ 
\end{bmatrix} x_1 + 
\begin{bmatrix} 
a_1 \\ 
... \\ 
a_t \\ 
\end{bmatrix}
$$
$$D_t \tilde{x}_t = \tilde{g}_t x_1 + \tilde{a}_t$$
$$D_t Var(\tilde{x}_t) D_t' = K^2 \tilde{g}_t \tilde{g}_t' + \tau^2 I_t$$
$$x_1 \sim N(0, K^2)$$
$$Var(\tilde{x}_t) = K^2 D_t^{-1} \tilde{g}_t \tilde{g}_t' D_t'^{-1} + \tau^2 D_t^{-1} D_t'^{-1} $$
$$E(\tilde{x}_t | \tilde{y}_t) = Var(\tilde{x}_t) \{ Var(\tilde{x}_t) + \sigma^2 I_t \}^{-1} \tilde{y}_t$$
$$Var(\tilde{x}_t | \tilde{y}_t) = Var(\tilde{x}_t) - Var(\tilde{x}_t) \{ Var(\tilde{x}_t) + \sigma^2 I_t \}^{-1} Var(\tilde{x}_t)$$
$$\implies \tilde{x}_t | \tilde{y}_t$$
$$\implies \tilde{y}_{t + k} | \tilde{y}_t$$
$$K = 0, \tau = 1, \sigma = 1 \implies Var(\tilde{x}_t) = D_t^{-1} D_t'^{-1} $$

```{r}
# Compute filter given all samples for each realization
D <- diag(n_t)
D[cbind(2:n_t, 1:(n_t - 1))] <- -1
var_x_t <- solve(D) %*% solve(t(D))
E_x_given_y <- var_x_t %*% solve(var_x_t + diag(n_t)) %*% y

# Plot it
matplot(t, y, type = 'l', main = "Random walk plus white noise", col = 1, 
        lty = 1)
quant_reps <- matrix(rep(c(0, 1.96, -1.96), each = n_samps), nrow = n_t)
filt_quants <- matrix(E_x_given_y, nrow = n_t, ncol = n_reals * 3)
matlines(t, filt_quants + quant_reps, col = 'blue', lty = rep(q_lty, each = n_reals))
grid()
# CI's look a bit wide but it's hard to tell, might be OK
```

